---
title: "Time series analysis with Quantile regression"
author: "Anton Antonov"
date: "10/1/2016"
output: html_document
---
<!---
##   Time series analysis with Quantile regression R-Markdown script
##   Copyright (C) 2016  Anton Antonov
##   
##   This program is free software: you can redistribute it and/or modify
##   it under the terms of the GNU General Public License as published by
##   the Free Software Foundation, either version 3 of the License, or
##   (at your option) any later version.
##   This program is distributed in the hope that it will be useful,
##   but WITHOUT ANY WARRANTY; without even the implied warranty of
##   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
##   GNU General Public License for more details.
##   You should have received a copy of the GNU General Public License
##   along with this program.  If not, see <http://www.gnu.org/licenses/>.
##   
##   Written by Anton Antonov,
##   antononcube @ gmail . com ,
##   Windermere, Florida, USA.
##============================================================
##   This R-Markdown file) is made for the R-part of the [MathematicaVsR](https://github.com/antononcube/MathematicaVsR/) project 
##     "Time series analysis with Quantile Regression"
##     https://github.com/antononcube/MathematicaVsR/tree/master/Projects/HandwrittenDigitsClassificationByMatrixFactorization).
##============================================================
--->

<!---
To run globally use:
rmarkdown::render( input = "./TimeSeriesAnalysisWithQuantileRegression.Rmd", output_file = "./TimeSeriesAnalysisWithQuantileRegression.html")
--->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document (R-Markdown file) is made for the R-part of the [MathematicaVsR](https://github.com/antononcube/MathematicaVsR/) project ["Time series analysis with Quantile Regression"](https://github.com/antononcube/MathematicaVsR/tree/master/Projects/HandwrittenDigitsClassificationByMatrixFactorization).

The main goal of this document is to demonstrate how to do in R:

- getting weather data (or other time series data),

- fitting Quantile Regression (QR) curves to time series data, and

- using QR to find outliers and conditional distributions.

## Libraries

```{r}
library(weatherData)
library(ggplot2)
library(reshape2)
library(quantreg)
library(splines)
```


## Getting time series data

Assume we want to obtain temperature time series data for Atlanta, Georgia, USA for the time interval from 2011.04.01 to 2016.03.31 .

Following the guide [2] we can download that weather data in the following way.

First we find weather stations identifiers in Atlanta, GA:
```{r}
getStationCode("Atlanta")
```

Let use the first one "KATL". The following code downloads the temperature data for desired time interval.

```{r get-time-series, message=FALSE}
if(!exists("tempDF")) {
  res <- 
    llply( seq(2011,2015), function(y) {
      getWeatherForDate( station_id = "KATL", 
                         start_date = paste(y, "04-01", sep="-" ), 
                         end_date = paste(y+1, "03-31", sep="-" ) )
    }, .progress = "None")
  tempDF <- do.call(rbind, res)
}
```

The obtained data frame has the following form:

```{r}
head(tempDF)
```

Below we are going to use the mean temperatures. Here is plot of that time series data:

```{r ts-ggplot, fig.height=4, fig.width=12}
ggplot(tempDF) + 
  geom_line(aes(x = Date, y = Mean_TemperatureF), color='dodgerblue3')
```
(The color name was selected from the web page ["ggplot2 Quick Reference: colour (and fill)"](http://sape.inf.usi.ch/quick-reference/ggplot2/colour).)

## Fitting Quantile regression curves and finding outliers

### QR fitting of B-splines

The package [`quantreg`](https://cran.r-project.org/web/packages/quantreg/index.html) provides several ways (functions and work flow) to apply QR to time series data. In this document we interested in applying QR using B-spline basis functions. Following the vignette [1] this can be done in the following way.

First we are going to add to the time series data frame an index column and an absolute time column.

```{r}
tempDF <- tempDF[order(tempDF$Date),]
tempDF <- cbind( tempDF, Index=1:nrow(tempDF), AbsTime = as.numeric(tempDF$Date) )
```

Next we make a model matrix for a selected number of knots.

```{r}
nKnots <- 30
X <- model.matrix( Mean_TemperatureF ~ bs(Index, df = nKnots + 3, degree = 3), data = tempDF )
```

We find the QR curves -- called regression quantiles -- at these quantiles:
```{r}
qs <- c(0.02,0.1,0.25,0.5,0.75,0.9,0.98)
```

Do the QR fit:
```{r qcurves}
qcurves <-
  llply( qs, function(x) {
    fit <- rq( Mean_TemperatureF ~ bs(Index, df = nKnots + 3, degree = 3), tau = x, data = tempDF)
    X %*% fit$coef
  }, .progress = "none")
```

We put the QR fitting result into a data frame with which further manipulations and plotting would be easier.
```{r}
qfitDF <- do.call(cbind, qcurves )
qfitDF <- data.frame(Index=1:nrow(qfitDF), Date = tempDF$Date, qfitDF )
```

### Finding outliers

At this point finding the outliers is simple -- we just pick the points (dates) with temperatures higher than the $0.98$ regression quantile (multiplied by some factor close to $1$, like, $1.005$.)
```{r}
outlierInds <- which( tempDF$Mean_TemperatureF > 1.005 * qfitDF[,ncol(qfitDF)] )
```

### Plot

The best way to plot the data is through melting into long form data frame. The identified outliers are given with red points.

```{r qcurves-outliers-ggplot, fig.height=4, fig.width=12 }
names(qfitDF) <- c( "Index", "Date", qs )
qfitMeltedDF <- melt( data = qfitDF, id.vars = .(Date, Index) )
names(qfitMeltedDF) <- gsub( "variable", "quantile", names(qfitMeltedDF) )

ggplot( tempDF ) + 
  geom_line( aes( x = Date, y = Mean_TemperatureF ), color = 'darkgrey' ) +
  geom_line( data = qfitMeltedDF, aes( x = Date, y = value, color = quantile ) ) +
  geom_point( data = tempDF[outlierInds, ], aes( x = Date, y = Mean_TemperatureF ), color = 'red')
```

## Re-construction of conditional probabilities distributions

### CDF and PDF re-construction function definitions

```{r}
CDFEstimateFunction <- function( qs, qvals ) {
  ## splinefun( x = qvals, y = qs, method = "natural" )
  approxfun( x = qvals, y = qs, method = "linear" )
}
```

Since we deal with piece-wise linear functions for CDF the PDF has to defined ad-hoc instead of using functions that find derivatives.

```{r}
PDFEstimateFunction <- function( qs, qvals ) {
  names(qvals) <- NULL; names(qs) <- NULL
  xs = ( qvals[-length(qvals)] + qvals[-1] ) / 2
  ys = diff(qs) / diff(qvals)
  approxfun( x = xs, y = ys, method = "constant" )  
}
```

Note, that if we used [`splinefun`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/splinefun.html) for the calculation of the CDF function `cdfFunc` we could implement the PDF function simply as `pdfFunc <- function(x) cdfFunc( x, 1 )`.

### QR with for lots of quantiles

Consider the quantiles:
```{r}
qs <- seq(0,1,0.05); qs <- c(0.02, qs[qs > 0 & qs < 1 ], 0.98); qs
```

With them we do following fitting (same code as above):
```{r}
qcurves <-
  llply( qs, function(x) {
    fit <- rq( Mean_TemperatureF ~ bs(Index, df = nKnots + 3, degree = 3), tau = x, data = tempDF)
    X %*% fit$coef
    }, .progress = "none")
qfitDF <- do.call(cbind, qcurves )
qfitDF <- data.frame(Index=1:nrow(qfitDF), Date = tempDF$Date, qfitDF )
```

### CDF and PDF re-construction

At this point we are ready to do the reconstruction of CDF and PDF for selected date and plot them.
```{r cdf-pdf-reconstr-ggplot, fig.height=3,fig.width=4}
ind <- 1100
qvals <- as.numeric(qfitDF[ind, 3:(2+length(qs))]); names(qvals) <- NULL
cdfFunc <- CDFEstimateFunction( qs, qvals )

xs <- seq(min(qvals),max(qvals),0.05)
print( 
  ggplot( ldply( xs, function(x) data.frame( X = x, CDF = cdfFunc(x), PDF = pdfFunc(x) ) ) ) + 
    geom_line( aes( x = X, y = CDF ), color = "blue" ) + 
    geom_line( aes( x = X, y = PDF ), color = "orange" ) + 
    ggtitle( paste( "CDF and PDF estimates for", qfitDF[ind, "Date"] ) ) +
    theme(plot.title = element_text(lineheight=.8, face="bold")) 
)
```

## References 

[1] Roger Koenker, ["Quantile regression in R: a vignette"](https://cran.r-project.org/web/packages/quantreg/vignettes/rq.pdf), (2015), [CRAN](https://cran.r-project.org/).

[2] Ram Narasimhan, ["weatherData: An R package that fetches Weather data from websites"](http://ram-n.github.io/weatherData/), http://ram-n.github.io/weatherData/.
