---
title: "Simple data reading and analysis functionalities"
author: "Anton Antonov"
date: "November 2016"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This document illustrates the use of base R functions for data reading and rudimentary data analysis.

## Data summary

Let us read a dataset provided with R's base system and print its summary:

```{r cars, echo = TRUE}
data("attitude")
summary(attitude)
```

In some cases we get not very useful results if some of the variables are with strings values:
```{r titanic, echo = TRUE}
library(titanic)
summary(titanic_train)
```

For these situations we can convert the string valued variables into factors and call `summary` again:

```{r, echo = TRUE}
summary(as.data.frame(unclass(titanic_train)))
```

Let us combine the training and testing data from the "titanic" package into one data frame:
```{r, echo = TRUE}
commonColNames <- intersect(names(titanic_train), names(titanic_test) )
titanicAll <- rbind( titanic_train[, commonColNames], titanic_test[, commonColNames])
cat("dim(titanicAll) = ", dim(titanicAll), "\n" )
summary(as.data.frame(unclass(titanicAll)))
```

We see that the resulting data frame does not have the column "Survival" because the data frame "titanic_test" does not have it.
Since we want to work with the survival data we are going to ignore `titanic_test` below. 

(In general, in R very often package creators take decisions that produce incomplete or unintuitive functionality and data. Those decisions are often undocumented or hard to understand.)

## Mosaic plots

Using a mosaic plot we can get (quickly) an impression of the co-dependencies of the categorical variables in a dataset.
We can get better looking results using `mosaic` of the package "vcd" instead of the base function `mosaicplot`.

```{r, echo = TRUE}
## mosaicplot( Survived ~ Sex + Pclass, titanic_train )
library(vcd)
mosaic( Survived ~ Sex + Pclass, titanic_train )
```

## Contingency values

Given the Titanic passengers data let us find the average age of the passengers in each combination of values of the variables "Sex" and "Survival".

1. First we find the passenger counts:

```{r, echo = TRUE}
pCountsDF <- xtabs( ~ Sex + Pclass, titanic_train )
pCountsDF
```

2. Then we find the total of passenger ages:

```{r, echo = TRUE}
pTotalAgesDF <- xtabs( Age ~ Sex + Pclass, titanic_train )
pTotalAgesDF
```

3. Finally we divide the total ages data frame by the counts data frame:

```{r, echo = TRUE}
pTotalAgesDF / pCountsDF
```

## Distributions plots

It is a good idea to get an impression of the numerical variables distributions in a given dataset.

There are several approaches for doing this (in R and in general.)

### Using base functions

First we can simply use the base function `boxplot`, e.g. `boxplot( attitude )`.
In the command below the function `boxplot` rotates and prints all column names by the argument `las=2` and uses alternating coloring by the argument `col=...`.

```{r, echo=TRUE}
boxplot( attitude, las=2, col = c("royalblue1","royalblue4") )
```

Alternatively, we can plot a histograms for all numerical columns.

### Using the package "lattice"

```{r, echo=TRUE}
library(lattice)
histogram( ~ values | ind, stack(attitude) )
```

### Using the package "ggplot2"

Let us get fencier. Note that is this a much harder to specify plot compared to the one made with "lattice" above.

```{r, echo=TRUE}
library(ggplot2)
ggplot(stack(attitude)) + aes( x = values, fill = ind ) + 
  geom_histogram( aes(y = ..density..), binwidth=10, alpha = 0.3) + 
  geom_density(alpha = 0,linetype=3) + 
  facet_grid( . ~ ind)
```


## Cross correllations plots

It useful to visualize scatter plots made by pairs of numerical variables in a dataset. 
This can be easily (readily) done with the function `pairs`:

```{r, echo = TRUE}
pairs(attitude)
```

Prettier and more informative plots can be obtained with other, non-base package functions.

### Using the package "GGally" based on "ggplot2"

```{r, echo=TRUE, fig.width=14, fig.height=12}
library(GGally)
ggpairs(attitude)
```

For more details and related plots see ["ggcorr: correlation matrixes with ggplot2"](https://briatte.github.io/ggcorr/).
