---
title: "Linear regression with ROC"
author: "Anton Antonov"
date: "10/10/2016"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

This document demonstrates how to do in R linear regression (easily using the built-in function `lm`) and to tune the binary classification with the derived model through the so called [Receiver Operating Characteristic](https://en.wikipedia.org/wiki/Receiver_operating_characteristic) (ROC) framework, [5, 6].

The data used in this document is from [1] and it has been analyzed in more detail in [2]. In this document we only show to how to ingest and do very basic analysis of that data before proceeding with the linear regression model and its tuning. The package ROCR, [3], (introduced with [4]) provides the needed ROC functionalities.

### Libraries needed to run the Rmd file:

```{r}
library(plyr)
library(ROCR)
library(lattice)
library(reshape2)
library(ggplot2)
```


## Data ingestion

The code below imports the data from [1].

```{r}
data <- read.table( "~/Datasets/adult/adult.data", sep = ",", stringsAsFactors = FALSE )  
testData <- read.table( "~/Datasets/adult/adult.test", fill = TRUE, sep = ",", stringsAsFactors = FALSE )  
testData <- testData[-1,]
testData[,1] <- as.numeric(testData[,1])

columnNames<-
  strsplit(paste0("age,workclass,fnlwgt,education,education.num,marital.status,occupation,",
                  "relationship,race,sex,capital.gain,capital.loss,hours.per.week,native.country,income"), ",")[[1]]

names(data) <- columnNames
names(testData) <- columnNames

data$income <- gsub( pattern = "\\s", replacement = "", data$income )
testData$income <- gsub( pattern = "\\s", replacement = "", testData$income )
testData$income <- gsub( pattern = ".", replacement = "", testData$income, fixed = TRUE )
```

## Assignment of training and tuning data

As usual in classification and regression problems we work with two data sets: a training data set and a testing data set. Here we split the original training set into two sets a training set and a tuning set. The tuning set is going to be used to find a good value of a tuning parameter through ROC.

```{r}
trainingInds <- sample( 1:nrow(data), ceiling( 0.8*nrow(data) ) )
tuningInds <- setdiff( 1:nrow(data), trainingInds )
trainingData <- data[ trainingInds, ]
tuningData <- data[ tuningInds, ]
```

## Basic data analysis

Before doing regression it is a good idea to do some preliminary analysis of the data. 

Here is the summary of the training data:
```{r}
summary(as.data.frame(unclass(data)))
```

And here is the summary of the test data:
```{r}
summary(as.data.frame(unclass(testData)))
```

For the code below we are going to use the following variables

```{r}
columnNameResponseVar <- "income"
columnNamesExplanatoryVars <- c("age", "education.num", "hours.per.week")
columnNamesForAnalysis <- c( columnNamesExplanatoryVars, columnNameResponseVar )
```

With this plot we can see that ```r columnNamesExplanatoryVars``` correlate (can explain) with ```r columnNameResponseVar```:

```{r}
dataLong <- melt( data = data[, columnNamesForAnalysis], id.vars = columnNameResponseVar  )
ggplot(dataLong, aes(x = income, y = value, fill = income)) + geom_violin() + facet_wrap( ~variable, ncol = 3)
```

On the plot above we see that higher values of ```r columnNamesExplanatoryVars``` are associated closer with ">50K". For more detailed analysis see [2].

## Linear regression

```{r}
dataReg <- trainingData[,columnNamesForAnalysis]
unique(dataReg$income)
dataReg$income <- ifelse( dataReg$income == ">50K", 1, 0 )

lmRes <- lm( income ~ age + education.num + hours.per.week, data = dataReg )  
```

## Linear regression with ROC

In this section we take a systematic approach of determining the best threshold to be used to separate the regression model values. 

We will consider ">50" to be the more important class label for the classifiers built below. As a result, we are going to call *positive* the income values ">50K" and *negative* the income values "<=50K".

The used ROC functionalities are employed through the package [3].

### Computations to find the best threshold

```{r}
modelValues <- predict(lmRes, newdata = tuningData[, columnNamesExplanatoryVars], type="response")

## unique(tuningData$income)

pr <- prediction( modelValues, ifelse( tuningData$income == ">50K", 1, 0) )
prf <- performance(pr, measure = "tpr", x.measure = "fpr")
ggplot( data.frame( FPR = prf@x.values[[1]], TPR = prf@y.values[[1]] ) ) + aes( x = FPR, y = TPR) + geom_line()
```

After looking at ```r str(prf)``` we can come up with the following code that plots the ROC functions "PPV", "NPV", "TPR", "ACC", and "SPC"/"SPEC".
```{r}
rocDF <- 
  ldply( c("ppv", "npv", "tpr", "acc", "spec"), function(x) {
    res <- performance(pr, measure = x, x.measure = "cutoff")
    data.frame( Measure = x, Cutoff = as.numeric(res@x.values[[1]]), Value = as.numeric(res@y.values[[1]]), stringsAsFactors = FALSE)
  })
rocDF <- rocDF[ !is.na(rocDF$Value), ]
ggplot(rocDF) + aes( x = Cutoff, y = Value, color = Measure) + geom_line() + coord_fixed(ratio = 1/1.2)
```

From the plot we can select the best cutoff value, in this case $\approx 0.3$.

### Accuracy over the test data

We split the original training data into two parts for training and tuning. Using the found threshold, let us use evaluate the classification process over the test data.

```{r}
modelValues <- predict(lmRes, newdata = testData[, columnNamesExplanatoryVars], type="response")

threshold <- 0.3
classDF <- data.frame( Actual = testData[, columnNameResponseVar], Predicted = ifelse( modelValues >= threshold, ">50K", "<=50K" ), stringsAsFactors = FALSE )
```

Here is the overall accuracy:
```{r}
mean( classDF$Actual == classDF$Predicted)
```

And here is the confusion matrix
```{r}
xtabs( ~ Actual + Predicted, classDF )
```
Here are the corresponding frequencies:
```{r}
xtabs( ~ Actual + Predicted, classDF ) / count( classDF, .(Actual))[,2]
```

## References

[1] Bache, K. & Lichman, M. (2013). [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml). Irvine, CA: University of California, School of Information and Computer Science. [Census Income Data Set](http://archive.ics.uci.edu/ml/datasets/Census+Income), URL: http://archive.ics.uci.edu/ml/datasets/Census+Income .

[2] Anton Antonov, "Classification and association rules for census income data", (2014), MathematicaForPrediction at WordPress.com , URL: https://mathematicaforprediction.wordpress.com/2014/03/30/classification-and-association-rules-for-census-income-data/ .

[3] [ROCR web site](http://rocr.bioinf.mpi-sb.mpg.de) [http://rocr.bioinf.mpi-sb.mpg.de](http://rocr.bioinf.mpi-sb.mpg.de).

[4] Tobias Sing, Oliver Sander, Niko Beerenwinkel, Thomas Lengauer. [ROCR: visualizing classifier performance in R](http://bioinformatics.oxfordjournals.org/cgi/content/abstract/21/20/3940), (2005), Bioinformatics 21(20):3940-3941.

[5] Wikipedia entry, Receiver operating characteristic. URL: http://en.wikipedia.org/wiki/Receiver_operating_characteristic .

[6] Tom Fawcett, An introduction to ROC analysis, (2006), Pattern Recognition Letters, 27, 861–874. ([Link to PDF](https://ccrma.stanford.edu/workshops/mir2009/references/ROCintro.pdf).)